---
title: "Untitled"
author: "Jami McMillen"
date: "2024-12-18"
output: pdf_document
---

## 1. Multiplication Script

### Description:
This section demonstrates the implementation of a multiplication table using both a "for" loop and a "while" loop. The goal is to display the multiplication table for numbers 1 through 10.

---
title: "Assignment-4.1"
author: "Jami McMillen"
date: "2024-12-18"
output: pdf_document
---

# 1. Multiplication Script

## Description:
This section demonstrates the implementation of a multiplication table using both a "for" loop and a "while" loop.

```{r, echo=TRUE}
# Using a for loop to generate a multiplication table
cat("Multiplication Table (For Loop):\n")
for (i in 1:10) {
  for (j in 1:10) {
    result <- i * j
    cat(i, "x", j, "=", result, "\n")
  }
  cat("\n")
}

cat("Multiplication Table (While Loop):\n")
i <- 1
while (i <= 10) {
  j <- 1
  while (j <= 10) {
    result <- i * j
    cat(i, "x", j, "=", result, "\n")
    j <- j + 1
  }
  cat("\n")
  i <- i + 1
}
```

2. Density Plot of Income
Description:
We will plot the density of the income variable from the custdata.tsv dataset. This plot will show the distribution of income, highlighting areas of concentration.

```{r}
# Load required packages
if (!require(ggplot2)) install.packages("ggplot2", repos = "http://cran.us.r-project.org")
library(ggplot2)

# Load customer data
custdata <- read.table("C:/Users/jamim/Downloads/custdata.tsv", header = TRUE, sep = "\t")

# Density plot
ggplot(custdata, aes(x = income)) +
  geom_density(fill = "skyblue", color = "blue", alpha = 0.6) +
  ggtitle("Density Plot of Income") +
  xlab("Income") +
  ylab("Density") +
  theme_minimal()

```

3. Bar Chart for Housing Type
Description:
In this section, a bar chart is created to show the distribution of housing types among customers, excluding the NA values.

```{r}
# Remove NA values in housing.type
custdata_clean <- subset(custdata, !is.na(housing.type))

# Plot bar chart
ggplot(custdata_clean, aes(x = housing.type)) +
  geom_bar(fill = "skyblue", color = "black") +
  ggtitle("Bar Chart of Housing Type") +
  xlab("Housing Type") +
  ylab("Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

4. Married Customers and Health Insurance
4.1: Extract Subset of Married Customers with Income > $50,000

```{r}
# Standardize health.ins to lowercase
custdata$health.ins <- tolower(as.character(custdata$health.ins))
married_high_income <- subset(custdata, marital.stat == "Married" & income > 50000)

# Display the subset
print(married_high_income)
```

4.2: Percentage of Married High-Income Customers with Health Insurance

```{r}
# Calculate percentage with health insurance
percentage_health_ins <- mean(married_high_income$health.ins == "true", na.rm = TRUE) * 100
cat("Percentage of married high-income customers with health insurance:", round(percentage_health_ins, 2), "%\n")
```

4.3: Compare with the Whole Dataset

```{r}
# Calculate overall percentage with health insurance
overall_percentage_health_ins <- mean(custdata$health.ins == "true", na.rm = TRUE) * 100
difference <- percentage_health_ins - overall_percentage_health_ins

cat("Overall percentage of customers with health insurance:", round(overall_percentage_health_ins, 2), "%\n")
cat("Difference in percentage:", round(difference, 2), "%\n")
```

5. Correlation Analysis

Description:
In this section, we calculate the correlation between the age, income, and num.vehicles variables from the custdata.tsv dataset, ensuring that we remove any invalid data points.

```{r}
# Clean data
custdata_clean <- subset(custdata, !is.na(income) & !is.na(num.vehicles) & !is.na(age))

# Calculate correlations
cor_age_income <- cor(custdata_clean$age, custdata_clean$income)
cor_income_vehicles <- cor(custdata_clean$income, custdata_clean$num.vehicles)
cor_age_vehicles <- cor(custdata_clean$age, custdata_clean$num.vehicles)

# Print results
cat("Correlation between Age and Income:", round(cor_age_income, 2), "\n")
cat("Correlation between Income and Number of Vehicles:", round(cor_income_vehicles, 2), "\n")
cat("Correlation between Age and Number of Vehicles:", round(cor_age_vehicles, 2), "\n")
```

6. Dating Dataset Analysis
6.1: Correlation Between Ice Cream, Miles, and Games

```{r}
# Load dating dataset
dating <- read.csv("C:/Users/jamim/Downloads/dating.csv")

# Calculate correlations
cor_icecream_games <- cor(dating$Icecream, dating$Games)
cor_miles_games <- cor(dating$Miles, dating$Games)

cat("Correlation between Ice Cream and Games:", round(cor_icecream_games, 2), "\n")
cat("Correlation between Miles and Games:", round(cor_miles_games, 2), "\n")
```

6.2: Regression of Miles to Predict Games

```{r}
# Regression
model <- lm(Games ~ Miles, data = dating)

# Regression plot
ggplot(dating, aes(x = Miles, y = Games)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  ggtitle("Regression of Games on Miles") +
  xlab("Miles") +
  ylab("Games") +
  theme_minimal()
```

6.3: Clustering Analysis Based on "Like"

```{r}
# Set up clustering
num_colors <- length(unique(dating$Like))
color_palette <- scales::hue_pal()(num_colors)

# Clustering plot
ggplot(dating, aes(x = Miles, y = Games, color = as.factor(Like))) +
  geom_point(size = 3, alpha = 0.6) +
  ggtitle("Scatter Plot Colored by Like") +
  xlab("Miles") +
  ylab("Games") +
  scale_color_manual(values = color_palette, name = "Like") +
  theme_minimal()
```

References:
Assistant. (2024). Discussion on R Markdown in R. OpenAI ChatGPT. Retrieved December 18, 2024, from https://chat.openai.com