---
title: "Predictive Analysis and Clustering of Washington State Housing Market Data in R"
author: "Jami McMillen"
date: "2024-12-20"
output: pdf_document
---

Data Importing and Reprocessing

```{r setup-import, include=TRUE, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(ggplot2)
library(caret)
library(corrplot)
library(dplyr)

# Import the dataset
file_path <- "C:/Users/jamim/Downloads/Dataset 2 (House Sales)/house_sales.csv"
housing_data <- read.csv(file_path)

# Check dimensions and structure
cat("Dimensions of the dataset:\n")
print(dim(housing_data))
cat("Structure of the dataset:\n")
str(housing_data)
```

Explanation

Loaded the dataset and inspected its structure and dimensions. The dataset contains 21 variables across 21,613 observations, with features ranging from numerical data - e.g., price, sqft_living - to categorical data - e.g., zipcode. This initial inspection helps us understand the scope of our analysis.




Data Cleansing and Feature Engineering

```{r}
# Handle missing data
cat("Summary of missing data:\n")
missing_data <- sapply(housing_data, function(x) sum(is.na(x)))
print(missing_data)

# Impute missing data for numerical columns using the median
housing_data <- housing_data %>%
  mutate(
    bedrooms = ifelse(is.na(bedrooms), median(bedrooms, na.rm = TRUE), bedrooms),
    bathrooms = ifelse(is.na(bathrooms), median(bathrooms, na.rm = TRUE), bathrooms),
    sqft_living = ifelse(is.na(sqft_living), median(sqft_living, na.rm = TRUE), sqft_living),
    sqft_lot = ifelse(is.na(sqft_lot), median(sqft_lot, na.rm = TRUE), sqft_lot)
  )

# Feature engineering
housing_data <- housing_data %>%
  mutate(
    house_age = 2024 - yr_built,
    total_sqft = sqft_living + sqft_lot,
    age_category = cut(yr_built, breaks = c(1900, 1950, 2000, 2022),
                       labels = c("Old", "Mid-Age", "New"))
  )

# Check for duplicates and remove them
housing_data <- distinct(housing_data)

# Save the cleaned dataset for reproducibility
write.csv(housing_data, "cleaned_housing_data.csv", row.names = FALSE)
```

Explanation

Missing values were imputed with the median for numerical columns. This ensures data completeness without introducing biases from extreme values.
Feature engineering included creating a house_age column and an age_category column to provide additional insights into the dataset.
Duplicates were removed to ensure data integrity.



Data Analysis and Visualization

## Categorical and numerical variable identification
```{r}
cat("Summary of categorical variables:\n")
print(sapply(housing_data, is.factor))
cat("Summary of numerical variables:\n")
print(sapply(housing_data, is.numeric))
```

## Summary statistics
```{r}
cat("Summary statistics:\n")
summary(housing_data)
```

Visualizations

## Price distribution
```{R}
ggplot(housing_data, aes(x = price)) +
  geom_histogram(binwidth = 100000, fill = "blue", color = "black") +
  ggtitle("Distribution of House Prices") +
  xlab("Price") +
  ylab("Frequency")
```

## Correlation heatmap
```{r}
numeric_data <- housing_data %>% select(where(is.numeric))
cor_matrix <- cor(numeric_data)
corrplot(cor_matrix, method = "circle", type = "upper", tl.col = "black", tl.srt = 45)
```

## Scatterplot for sqft_living vs price
```{r}
ggplot(housing_data, aes(x = sqft_living, y = price)) +
  geom_point(color = "blue") +
  ggtitle("Living Area vs Price")
```
Interpretation

Price Distribution - The price distribution is right-skewed, indicating a majority of houses are priced below the mean price, with fewer high-value properties.
Correlation Heatmap - Strong positive correlations were observed between sqft_living and price. This suggests larger houses tend to be more expensive.
Scatterplot - The positive trend between sqft_living and price confirms the correlation and highlights the importance of living area as a predictor for price.



# Boxplot of price by age category
```{r}
ggplot(housing_data, aes(x = age_category, y = price, fill = age_category)) +
  geom_boxplot() +
  ggtitle("Price Distribution by Age Category") +
  xlab("Age Category") +
  ylab("Price")
```

# Data Analytics: Regression Analysis
```{r}
# Split the data into training and test sets
set.seed(123)
train_index <- createDataPartition(housing_data$price, p = 0.8, list = FALSE)
train_data <- housing_data[train_index, ]
test_data <- housing_data[-train_index, ]

# Linear regression
lm_model <- lm(price ~ sqft_living + bedrooms + bathrooms + sqft_lot, data = train_data)
summary(lm_model)
```

# Predictions and MSE
```{r}
predictions <- predict(lm_model, test_data)
mse <- mean((test_data$price - predictions)^2)
cat("Mean Squared Error for Linear Regression:", mse, "\n")
```
Interpretation

Regression Model - The model explains 46.87% of the variance in house prices - R-squared 0.4687. sqft_living has the highest positive impact, while bedrooms negatively affect price.
Mean Squared Error (MSE) - An MSE of roughly 73.6 billion indicates the model has reasonable predictive accuracy but can be improved with additional features.



# Clustering Analysis
```{r}
# Scale the numerical features for clustering
scaled_data <- scale(select(housing_data, c(sqft_living, sqft_lot, bathrooms, bedrooms)))

# K-means clustering
set.seed(123)
kmeans_model <- kmeans(scaled_data, centers = 3)

# Add cluster labels to the dataset
housing_data$cluster <- factor(kmeans_model$cluster)

# Visualize clusters
ggplot(housing_data, aes(x = sqft_living, y = sqft_lot, color = cluster)) +
  geom_point() +
  ggtitle("K-means Clustering of Houses")
```
Interpretation

Clustering - Three clusters represent distinct groupings based on house characteristics, sqft_living, sqft_lot, etc.
Insights - Cluster 1 includes small houses with small lots, while Cluster 3 includes large houses with larger lots. Clustering helps identify market segments and potential areas for targeted pricing strategies.










References
OpenAI. (2024). ChatGPT response on R Markdown refinement. Personal communication.